Computational methods in landscape ecology

### 2024-09-16 Sections/03-connectivity.qmd
One sentence per line

<!-- Person responsible: Hesselbarth, Gerber?, Simpkins? -->

@@

Landscape connectivity is a fundamental concept in spatial ecology. It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity]. Landscape connectivity can be divided into two fundamental types, structural and functional. Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency. It is solely a landscape feature and is independent of any particular species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity]. Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of a particular species, or set of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity]. An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of optimal habitat for a species.

@@

## Estimating connectivity
While connectivity is key to many modern conservation theories and practices, directly quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity). Data from many individuals recording the start location, trajectory, and endpoint of a movement is required to directly observe connectivity [@Simpkins2018AssessingPerformance]. Using this data one may be able to infer the elements in the landscape that individuals preferentially move through. While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution. As such in practice connectivity is primarily determined by indirect estimation. @Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
- Structural connectivity: determined by the physical attributes of the landscape (as described above).
- Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscpae elements or habitat patches.

@@

#### Least-cost paths (LCP)
LCP is among the most widely used methods for quantifying functional connectivity. The underlying assumption is that organisms will move through landscapes in ways that minimize the cumulative cost of traversing different landscape elements. LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effient than others

>>>>

<!-- Person responsible: Gerber?, Simpkins? -->

@@

Landscape connectivity is a fundamental concept in spatial ecology. 
It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity]. 
Landscape connectivity can be divided into two fundamental types, structural and functional. 
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency. 
It is solely a landscape feature and is independent of any particular species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity]. 
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of a particular species, or set of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity]. 
An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of optimal habitat for a species.

@@

## Estimating connectivity
While connectivity is key to many modern conservation theories and practices, directly quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity). 
Data from many individuals recording the start location, trajectory, and endpoint of a movement is required to directly observe connectivity [@Simpkins2018AssessingPerformance]. 
Using this data one may be able to infer the elements in the landscape that individuals preferentially move through. 
While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution. 
As such in practice connectivity is primarily determined by indirect estimation. @Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
- Structural connectivity: determined by the physical attributes of the landscape (as described above).
- Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscape elements or habitat patches.

@@

#### Least-cost paths (LCP)
LCP is among the most widely used methods for quantifying functional connectivity. 
The underlying assumption is that organisms will move through landscapes in ways that minimize the cumulative cost of traversing different landscape elements. 
LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. 
The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effcient than others
